---
title: "Nesting Data Manipulation"
author: "Justin Clarke"
date: "8/27/2021"
output: html_document
---

First I need to read in the raw data and calculate the average robel readings for later in my analysis.
All nesting data has used the midpoint of the last monitoring period as the date of fledging or failure unless evidence shows otherwise.

```{r}
setwd("~/Git/NDSU/AvianCommunityAnalysis")                                                             # setting the working directory

library(lubridate)
library(vegan)
library(tidyverse)

raw <- read.csv("~/Git/NDSU/AvianCommunityAnalysis/RawData/Nesting.csv")                                                  # read in the data set

raw$Date <- as.Date(raw$Date,                                                   # read the Date column as a date
                      "%m/%d/%y") |>                                            # show the format of the existing dates
    yday()                                                                      # calculate the Julian day

raw$aRobel <- rowMeans(raw[,28:31])                                             # create a new column with the robel readings averaged

# create a column for each data frame with the associated group information
SpeciesGrouping <- function(x) {
x <- x |>                                                                       # select the data frame
  mutate(Group = case_when(                                                     # group together each species and
    endsWith(Spec, "STGR") ~ "OBL",                                             # list each species as FAC or OBL based on Vickery et al. 1999
    endsWith(Spec, "UPSA") ~ "OBL",
    endsWith(Spec, "SAVS") ~ "OBL",
    endsWith(Spec, "GRSP") ~ "OBL",
    endsWith(Spec, "CCLO") ~ "OBL",
    endsWith(Spec, "DICK") ~ "OBL",
    endsWith(Spec, "BOBO") ~ "OBL",
    endsWith(Spec, "WEME") ~ "OBL",
    endsWith(Spec, "GADW") ~ "FAC",
    endsWith(Spec, "AMWI") ~ "FAC",
    endsWith(Spec, "MALL") ~ "FAC",
    endsWith(Spec, "BWTE") ~ "FAC",
    endsWith(Spec, "NSHO") ~ "FAC",
    endsWith(Spec, "NOPI") ~ "FAC",
    endsWith(Spec, "GWTE") ~ "FAC",
    endsWith(Spec, "WILL") ~ "FAC",
    endsWith(Spec, "MODO") ~ "FAC",
    endsWith(Spec, "EAKI") ~ "FAC",
    endsWith(Spec, "CCSP") ~ "FAC",
    endsWith(Spec, "RWBL") ~ "FAC",
    endsWith(Spec, "BRBL") ~ "FAC",
    endsWith(Spec, "KILL") ~ "GEN", 
    endsWith(Spec, "WIPH") ~ "GEN", 
    endsWith(Spec, "WISN") ~ "GEN", 
    endsWith(Spec, "YHBL") ~ "GEN",
    endsWith(Spec, "COGR") ~ "GEN",
    endsWith(Spec, "PBGR") ~ "GEN",
    endsWith(Spec, "SORA") ~ "GEN",
    endsWith(Spec, "AMCO") ~ "GEN",
    endsWith(Spec, "MAWR") ~ "GEN"
                           ))
}

raw <- SpeciesGrouping(raw) |> 
  drop_na(Group)
```

I need to shorten the data to just show each species rather than each observation (multiple of each species). I also need to assign each of those individuals an abundance so that I can total them and determine how many of each species is in each pasture/treatment

```{r}
unique <- raw |>                                                                # select the data frame
  group_by(id,                                                                  # group data by Nest.ID in order to remove duplicate visits
           Spec,                                                                # species
           Pasture,                                                             # pasture
           cTreat,                                                              # current treatment
           Patch,
           Year,
           Group) |>                                                            # treatment
  summarize()                                                                   # summarize the data to only show the grouped variables

unique <- subset(unique,                                                        # this selects only unique observations (i.e. each individual).
                 Spec!="DUCK" & cTreat!= "")                                    # it also removes any columns with no id and/or no treatment info
 
# this is the totals for the 2021 data grouped by pasture, species, and current treatment
totals21 <- unique |>                                                           # select the dataframe
  filter(Year=="2021") |>                                                       # filter the 2021 data out
  group_by(Pasture,                                                             # group by pasture
           Spec,                                                                # species
           cTreat,                                                              # current treatment
           Group) |>                                                            # species group (i.e. OBL, FAC, Gen)
  summarize(Abundance=n()) |>                                                   # count all abundances and create a new abundance column
  as.data.frame()                                                               # convert to a data frame

# this is the totals for the 2021 data grouped by pasture, species, and current treatment
totals22 <- unique |>                                                           # select the dataframe
  filter(Year=="2022") |>                                                       # filter the 2021 data out
  group_by(Pasture,                                                             # group by pasture
           Spec,                                                                # species
           cTreat,                                                              # current treatment
           Group) |>                                                            # species group (i.e. OBL, FAC, Gen)
  summarize(Abundance=n()) |>                                                   # count all abundances and create a new abundance column
  as.data.frame()                                                               # convert to a data frame

patch <- unique |> 
  group_by(Patch,
           Spec,
           cTreat,
           Group) |> 
  summarize(Abundance=n()) |> 
  as.data.frame()

totals <- patch |> 
  group_by(Spec,
           cTreat,
           Group) |> 
  summarize(Abundance=sum(Abundance)) |> 
  as.data.frame()

totals.mat <- pivot_wider(totals[c(1:2,4)],                                           # select the data frame to turn into a matrix
                        names_from = cTreat,                                      # select the column names
                        values_from = Abundance,                                # select the abundance values for each species
                        values_fill = list(Abundance = 0)) |>                   # fill all NA with 0
  column_to_rownames("Spec") |>
  relocate(Heavy,
           .before = Full)

totals.mat$SpecTotals <- rowSums(totals.mat)
totals.mat <- rbind(totals.mat, "Totals"= colSums(totals.mat))

totals.mat <- rownames_to_column(totals.mat, var = "Spec")

totals.mat <- SpeciesGrouping(totals.mat)

totals.mat <- column_to_rownames(totals.mat, var = "Spec")

write.csv(raw, "~/Git/NDSU/AvianCommunityAnalysis/WorkingData/RAWadjusted.csv")
write.csv(totals.mat, "~/Git/NDSU/AvianCommunityAnalysis/WorkingData/totals.csv")
write.csv(totals21, "~/Git/NDSU/AvianCommunityAnalysis/WorkingData/totals21.csv")
write.csv(totals22, "~/Git/NDSU/AvianCommunityAnalysis/WorkingData/totals22.csv")
```
